---
title: Multiple Linear Regression and Maximum Likelihood Estimation
---

# Multiple Linear Regression
```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(patchwork)
library(broom)
library(modelr)
```
For this example, we explore the power of the log transformation, along with a downside.  
We start by looking at air quality data from New York City, May to September 1973.  
The response of interest here is the Ozone, so we also add the log transformation in the dataset.
Our predictors will be solar.R, wind, and temp.  So the models we will look at are
$$
y_i = \beta_0 + \beta_1x_{1,i} + \beta_2x_{2,i} + \beta_3x_{2,i} + \epsilon,
$$
and 
$$
\log y_i = \beta_0 + \beta_1x_{1,i} + \beta_2x_{2,i} + \beta_3x_{2,i} + \epsilon.
$$
```{r}
air_quility_df = airquality |>
  as_tibble() |>
  mutate(log_ozone = log(Ozone)) |>
  na.omit()
glimpse(air_quility_df)
```

When doing regression is it is always good to start by creating plots of our 
predictors with the response.  This results in the following for the natural scale
Ozone (first row) and log scale Ozone (second row).
```{r}
#| echo: false
#| message: false
#| warning: false
solar_ozone_plot = ggplot(data = air_quility_df, aes(x = Solar.R, y = Ozone)) +
  geom_point() +
  theme_bw()
wind_ozone_plot = ggplot(data = air_quility_df, aes(x = Wind, y = Ozone)) +
  geom_point() +
  theme_bw()
temp_ozone_plot = ggplot(data = air_quility_df, aes(x = Temp, y = Ozone)) +
  geom_point() +
  theme_bw()
solar_log_ozone_plot = ggplot(data = air_quility_df, aes(x = Solar.R, y = log_ozone)) +
  geom_point() +
  theme_bw()
wind_log_ozone_plot = ggplot(data = air_quility_df, aes(x = Wind, y = log_ozone)) +
  geom_point() +
  theme_bw()
temp_log_ozone_plot = ggplot(data = air_quility_df, aes(x = Temp, y = log_ozone)) +
  geom_point() +
  theme_bw()

(solar_ozone_plot + wind_ozone_plot + temp_ozone_plot) /
  (solar_log_ozone_plot + wind_log_ozone_plot + temp_log_ozone_plot)
```

### Question :
What can you notice when be compare the log and natural scaled Ozone plots?  Which appears to be more linear?

## Fitting MLR to Both :
```{r}
model_no_log = lm(Ozone ~ Solar.R + Wind + Temp, data = air_quility_df)
model_log = lm(log_ozone ~ Solar.R + Wind + Temp, data = air_quility_df)
```

Lets look at what our estimates are for regression coefficients with corresponding
p-values.

```{r}
tidy(model_no_log)
tidy(model_log)
```

In our log model, what is the notation that would represent 0.0492?


## Checking Conditions for Estimates
1. Linearity
   a. addressed above
2. Normality of residuals
3. Constant Variance
4. Independence

### Question :
Which conditions does the natural scale Ozone model violate?  Which conditions 
does the log scale Ozone model violate?  *Use the plots below.*
```{r}
#| echo: false
#| message: false
#| warning: false
residuals_no_log_df = tibble(
  fitted_values = fitted.values(model_no_log),
  residuals = residuals(model_no_log)
)
residuals_log_df = tibble(
  fitted_values = fitted.values(model_log),
  residuals = residuals(model_log)
)
res_dens_no_log_plot = ggplot(residuals_no_log_df, aes(x = residuals)) +
  geom_density(color = "dodgerblue", fill = "dodgerblue", alpha = .1) +
  labs(title = "Distribution of residuals",
       subtitle = "for Model with Ozone on natural scale.") +
  theme_bw()
res_dens_log_plot = ggplot(residuals_log_df, aes(x = residuals)) +
  geom_density(color = "dodgerblue", fill = "dodgerblue", alpha = .1) +
  labs(title = "Distribution of residuals",
       subtitle = "for Model with Ozone on log scale.") +
  theme_bw()
res_no_log_plot = ggplot(residuals_no_log_df, aes(x = fitted_values, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "dodgerblue") +
  labs(title = "Residuals vs Fitted Values", 
       subtitle = "for Model with Ozone on natural scale.") +
  theme_bw()
res_log_plot = ggplot(residuals_log_df, aes(x = fitted_values, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "dodgerblue") +
  labs(title = "Residuals vs Fitted Values", 
       subtitle = "for Model with Ozone on log scale.") +
  theme_bw()

(res_dens_no_log_plot + res_dens_log_plot) /
  (res_no_log_plot + res_log_plot)
```

## Interpretation
When interpreting coefficients for a model, we must remember to convert transformed
variables back into their original scale.  So, in the log scale, we need to use
$$
\hat{y_i} = \exp\left[ b_0 + b_1x_{1,i} + b_2x_{2,i} + b_3x_{2,i}\right]
$$
as a basis for interpreting coefficients.

### Question 
Give an interpretation for $b_3$ in both the natural scale and log scale models.


Using notation, write out the hypotheses to test for the slope coefficient for wind. Then, using the output from the log scale model, determine the results of this hypothesis test.

```{r}
confint(model_no_log)
```

Interpret the 95% CI for the Wind slope coefficient.


# Maximum Likelihood Estimation (MLE)

## Steps of MLE
1. Find the log-likelihood function, $\log L(\theta; x_1,...,x_n)$.
2. Take the derivative of the log-likelihood with respect to the parameter we are estimating, $\frac{d}{d\theta}\log L(\theta;x_1,...,x_n)$.
3. Set derivative to zero and solve for parameter we are interested in estimating, $\frac{d}{d\theta}\log L(\theta;x_1,...,x_n) = 0$.
4. Check the second derivative for concavity (negative), we want a maximum, $\frac{d^2}{d\theta^2}\log L(\theta;x_1,...,x_n)$.


## You Try! Normal level
Suppose we have independently identically distribution random variables, $X_i$, where $X_i \sim \text{Poisson}(\lambda)$. Find the MLE of $\lambda$. Recall that the pmf of this random variable is

$$
f(X_i|\lambda) = \frac{\lambda^{x_i} e^{-\lambda}}{x_i!}
$$



## You Try! Difficult Level
Suppose we have independent identically distributed random variables, $Y_i$, where $Y_i \sim \text{Normal}(x_i\beta, 1)$, where $x_i$'s are given for $i = 1,...,n$.  Show that the MLE of $\beta$ is 
$\frac{\sum_{i=1}^{n}x_iY_i}{\sum_{i=1}^{n}x_i^2}$.  Recall that the pdf of this random variable is
$$
f(Y_i|x_i,\beta) = \frac{1}{\sqrt{2\pi}}\exp\left[-\frac{(Y_i - x_i\beta)^2}{2}\right]
$$


# Final Review (textbook problems):

Paired vs non-paired data: 7.15, 7.17, 7.18

Linear regression: 8.25, 8.26, 8.31(a-c), 8.32(a-c), 8.33, 8.35

1-prop: 6.8, 6.10, 6.45, 6.48

2-prop: 6.46, 6.17, 6.18, 6.21 (and conduct a hypothesis test if you can), 6.22, 6.24

1-mean: 7.1, 7.3, and 7.4 (write R code); 7.7, 7.11, 7.12, 7.31 (try to identify the variables), 7.51, 7.56; 7.57 & 7.58 (you can use this to practice writing hypotheses)

2-mean: 7.47, 7.48, 7.49,  



Since there are **no calculators**, we can expect: 

- R-code questions may include calculating the p-value, calculating the critical value for a CI, etc.

- identifying hypotheses

- identifying sampling distributions

- identifying conditions for inference (independence, etc.)

- interpreting CI, p-value, and hypothesis test results

- there will be at least one question on understanding studies and identifying their variables, drawing conclusions, etc. (1.41, 1.42)

