# Load Packages

```{r}
library(datasets)
library(tidyverse)
```


# About R

Basic functions and language

```{r}
print('hello woRld')
birth_years <- c(1996, 2001, 1999)
print(birth_years)
```

# Dataframe in R

```{r}
data.frame(person = c('Jack', 'Allen', 'Lisa'), birth_years = birth_years)

?data.frame
```

You could also check the use of a function or a package by searching in the help panel.

# Tidyverse

After function names do not leave any spaces.

Before and after operators (e.g. <-, =) leave spaces.

Put a space after a comma, not before.

Object names are all lower case, with words separated by an underscore.


# Markdown and Quarto

word

_word_

__word__

~~word~~

If you need R variables in text, you can do this: the birth_years are:
`r birth_years`

# Load mtcars Dataset

```{r}
mtcars <- mtcars %>% 
  mutate(am = as.factor(am))
```

# Dataset

## Description given by the author

```{r}
?mtcars
```

# First Look

Take a glimpse at the data.

1.  How many rows and columns are there in the "mtcars" dataset?

There are 32 rows and 11 columns

2.  Any categorical variables?

'vs' and 'am' are categorical variables (both are binary).

3.  Any numeric variables?

The other 9 variables are numeric.

```{r}
glimpse(mtcars)
```

# First and last few lines

```{r}
head(mtcars)
```

```{r}
tail(mtcars)
```


# Summarizing Categorical Data

## Count by Transmission Types (am)

How many samples are there in each transmission type?

```{r}
count(mtcars, am)
```

## Bar plots

### Bar plot with counts

```{r}
ggplot(mtcars) +
  aes(x = am) +
  geom_bar()
  
```

# Summarizing Numeric Data

## A few basic statistics

### Mean

What is the meaning of "Mean"?

The mean is the average value which is calculated by summing all of the observations and dividing by the total number of observations.

```{r}
summarize(mtcars, mean(mpg))
```

### Median

What is the meaning of "Median"?

The median is another measure of center that is determined by choosing the middle value when the observations are ordered from smallest to largest.

```{r}
summarize(mtcars, median(mpg))
```

### Quantiles

The 3rd Quantile of mpg:

```{r}
summarize(mtcars, quantile(mpg, 0.75))
```

What about the 1st Quantile? What is the meaning of "Quantile"?

```{r}
summarize(mtcars, quantile(mpg, 0.25))
```

Quantiles divide the data into quarter. Thus, 25\% of the data is less than the 1st quantile while 75\% of the data is less than the 3rd quantile.

### Standard Deviation and Variance

What are the meanings of "Standard Deviation" and "Variance"?

Standard deviation and variance are both measures of the spread of the data. See the Week 1 lecture slides for the formulas to calculate standard deviation and variance.

```{r}
summarize(mtcars, sd(mpg))
```

```{r}
summarize(mtcars, var(mpg))
```

## Histogram

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 8)
```

## Boxplot of Miles per Gallon (mpg)

On the boxplot, where is the median? Q1 and Q3? Interquartile Range (IQR)? Whiskers?

```{r}
ggplot(mtcars,
       aes(y = mpg)) + 
  scale_x_discrete( ) +
  geom_boxplot()
```

The median is the middle line of the box. Q1 and Q3 are the bottom and top lines of the box. The IQR is Q3 - Q1. Whiskers help show the spread of the data.

# Summarizing More Than One Variable

## Boxplot of Miles per Gallon (mpg) by Transmission Types (am)

How many variables are represented in this plot?

Are they numeric or categorical?

```{r}
ggplot(mtcars, aes(x = am, y = mpg)) +
  geom_boxplot()
```
The plot represents two variables ('am' and 'mpg'). 'am' is categorical while 'mpg' is numeric.

## Scatterplot of Miles per Gallon (mpg) and Horse Power (hp)

How many variables are represented in this plot?

Are they numeric or categorical?

```{r}
ggplot(mtcars, aes(x = mpg,
                 y = hp)) +
  geom_point()
```
We now plot two numeric variables ('mpg' and 'hp')



