---
title: "Discussion 2: Summarizing Data & Study Design"
author: "Stats 67 Teaching Team"
format: html
---


## 1. Summarizing Data

## Load Packages

```{r}
library(datasets)
library(tidyverse)
```

### Load mtcars Dataset

```{r}
mtcars <- mtcars %>% 
  mutate(am = as.factor(am))
```

## Dataset

### Description given by the author

```{r}
?mtcars
```

## Summarizing Categorical Data

### Count by Transmission Types (am)

How many samples are there in each transmission type?

```{r}
count(mtcars, am)
```

### Bar plots

#### Bar plot with counts

```{r}
ggplot(mtcars) +
  aes(x = am) +
  geom_bar()
  
```

## Summarizing Numeric Data

### A few basic statistics

#### Mean

What is the meaning of "Mean"?

```{r}
summarize(mtcars, mean(mpg))
```

#### Median

What is the meaning of "Median"?

```{r}
summarize(mtcars, median(mpg))
```

#### Quantiles

The 1st Quantile of mpg:

```{r}
summarize(mtcars, quantile(mpg, 0.25))
```

What about the 1st Quantile? What is the meaning of "Quantile"?

#### Standard Deviation and Variance

What are the meanings of "Standard Deviation" and "Variance"?

```{r}
summarize(mtcars, sd(mpg))
```

```{r}
summarize(mtcars, var(mpg))
```

---

### Histogram

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 8)
```

### Boxplot of Miles per Gallon (mpg)

On the boxplot, where is the median? Q1 and Q3? Interquartile Range (IQR)? Whiskers?

```{r}
ggplot(mtcars,
       aes(y = mpg)) + 
  scale_x_discrete( ) +
  geom_boxplot()
```

## Summarizing More Than One Variable

### Boxplot of Miles per Gallon (mpg) by Transmission Types (am)

How many variables are represented in this plot?

Are they numeric or categorical?

```{r}
ggplot(mtcars, aes(x = am, y = mpg)) +
  geom_boxplot()
```

---

### Scatterplot of Miles per Gallon (mpg) and Horse Power (hp)

How many variables are represented in this plot?

Are they numeric or categorical?

```{r}
ggplot(mtcars, aes(x = mpg,
                 y = hp)) +
  geom_point()
```



## 2. Sampling

Sampling is the process of selecting a subset of individuals or items from a larger population for the purpose of drawing conclusions about the population.  Let's dive into random sampling and explore some real examples.

---

### Random Sampling

Random sampling ensures each individual/item in the population has an equal chance of being selected. It reduces bias in your sample.

- **What is the difference between a population and a sample?**  

---

**Example**: Let's consider a new example for sampling. You want to survey the average number of hours university students spend on social media per day.

- **What is the population in this scenario?**
- **What is the sampling method you would select for this survey?** Do not just write the title of the sampling method but also explain the steps you are utilizing (e.g., get a list of registrar or post on a website etc).
- **Is your sample a random sample? Why or why not?**
- **Could there be any bias in the sampling process? What kind of bias might exist?**

---

### Random Sampling vs. Convenience Sampling

Read the following two scenarios and decide which one uses random sampling and which uses convenience sampling.

1. You randomly select 200 students from the entire student population at your university.
2. You survey students waiting in line at a campus coffee shop.

**Which sampling method is less likely to introduce bias? Why?**

---

## 3. Study Design in Research

The way a study is designed plays a critical role in the validity and reliability of the conclusions drawn from it. Let’s explore **observational studies** and **experimental designs**.

---

### Observational Study

In an observational study, researchers observe the subjects without interfering. There is no treatment or intervention involved.
 
- **What are the strengths of an observational study?**  
- **What are the limitations?**

---

### Experiment Design

An experimental design involves manipulating one or more variables (treatments) and observing the effects.

**Example**: Suppose you are studying the effect of two different diets on weight loss over a 6-week period. The two diets are a low-carb diet and a low-fat diet.

- **What is the treatment in this experiment?**
- **What variables will you control for (e.g., exercise, initial weight)?**

