---
title: "Discussion 2: Study Design & Intro to Probability"
author: "Stats 67 Teaching Team"
format: html
---

## 1. Sampling

Sampling is the process of selecting a subset of individuals or items from a larger population for the purpose of drawing conclusions about the population.  Let's dive into random sampling and explore some real examples.

---

### Random Sampling

Random sampling ensures each individual/item in the population has an equal chance of being selected. It reduces bias in your sample.

- **What is the difference between a population and a sample?**  

A population is the entire group that your research question pertains to. A sample is a subset of the population that you study and collect data on.

---

**Example**: Let's consider a new example for sampling. You want to survey the average number of hours university students spend on social media per day.

- **What is the population in this scenario?**

The population is all university students.

- **What is the sampling method you would select for this survey?** Do not just consider the title of the sampling method but also explain the steps you are utilizing (e.g., get a list of registrar or post on a website etc).

There are many sampling methods with different advantages and disadvantages. Good sampling methods would limit bias by surveying students in a random manner.

- **Is your sample a random sample? Why or why not?**

Depending on the sampling method, we could have a random sample.

- **Could there be any bias in the sampling process? What kind of bias might exist?**

There could be bias in the sampling process, although the amount varies based on the sampling method. Since data collection is likely to rely on voluntary responses by students, we are likely to see some inherent bias based on which students actually choose to respond to surveys.

---

### Random Sampling vs. Convenience Sampling

Read the following two scenarios and decide which one uses random sampling and which uses convenience sampling.

1. You randomly select 200 students from the entire student population at your university.
2. You survey students waiting in line at a campus coffee shop.

**Which sampling method is less likely to introduce bias? Why?**

Option 1 is less likely to introduce bias. Option 2 inherently restricts you to students who go to this specific campus coffee shop at the particular time that you choose to survey them.

---

## 2. Study Design in Research

The way a study is designed plays a critical role in the validity and reliability of the conclusions drawn from it. Let’s explore **observational studies** and **experimental designs**.

---

### Observational Study

In an observational study, researchers observe the subjects without interfering. There is no treatment or intervention involved.
 
- **What are the strengths of an observational study?**  
- **What are the limitations?**

Observational studies are typically easier and cheaper methods of collecting data. However, because researchers do not have control over sampling method or data collection, there is the potential for bias or confounders.

---

### Experiment Design

An experimental design involves manipulating one or more variables (treatments) and observing the effects.

**Example**: Suppose you are studying the effect of two different diets on weight loss over a 6-week period. The two diets are a low-carb diet and a low-fat diet.

- **What is the treatment in this experiment?**

The treatment is the type of diet.

- **What variables will you control for (e.g., exercise, initial weight)?**

Variables that could otherwise have an impact on weight loss (exercise, initial weight, etc.) should be controlled for.

### Simpson's Paradox

Simpson's Paradox is a statistical phenomenon that can lead to counterintuitive results when analyzing data. 
We'll explore Simpson's Paradox using an example involving Kidney stone treatment.

Example: Kidney stone treatment


```{r}
total_patients = c(87, 263, 270, 80)
success = c(81, 191, 234, 55)
# Example data
data <- data.frame(
  treatment = c("A", "A", "B", "B"),
  stone = c("S", "L", "S", "L"),
  total_patients = c(87, 263, 270, 80),
  success = c(81, 191, 234, 55)
)

# Calculate mortality rates by hospital and surgery type
success_rates <- aggregate(success / total_patients, by = list(data$treatment, data$stone), FUN = mean)

success_rates
```

It appears that Treatment A has a higher success rate for both kind of stones, while treatment B has a lower success rate for them.

However, when we aggregate the data without considering the stone type:

The overall success rates for each treatment are as follows:

| Treatment| Success Rate   |
|----------|----------------|
| A        | 78%            |
| B        | 83%            |

Surprisingly, when we look at the overall success rates without considering the stone type, treatment B seems better. This is an example of Simpson's Paradox.

Implications:

Simpson's Paradox is a reminder that drawing conclusions from aggregated data without considering all relevant factors can lead to incorrect or misleading insights. When encountering this paradox, it's essential to dig deeper into the data, consider confounding variables, and analyze subgroups to gain a more accurate understanding of the relationships within the data.


## 3. Intro to Probability

Probability helps us quantify uncertainty and make predictions about the likelihood of events.

---

### Sample Space and Events

- **Sample Space (S)**: The set of all possible outcomes of an experiment.
- **Event (E)**: A specific outcome or set of outcomes within the sample space.

**Think About This**

- **If you flip a fair coin, what is the sample space?**
- **If your event is getting heads, what is the outcome for this event?**

```{r}
# Simulate the sample space for flipping a coin
sample_space <- c("Heads", "Tails")

# Event: Getting Heads
event <- sample_space[1]

# Display results
sample_space
event
```


---

### Defining Probability

- **Probability (P)**: The likelihood that an event will occur, ranging from 0 (impossible) to 1 (certain).
  
**Example**:  
If you flip a fair coin, there are two possible outcomes: heads or tails. Each outcome has a probability of 0.5.

**Try It**:

- **What is the probability of drawing a face card from a standard deck of 52 cards?**
- **What is the probability of rolling an odd number on a fair six-sided die?**

```{r}
# Probability of drawing a face card (12 face cards in a deck of 52)
prob_face_card <- 12 / 52

# Probability of rolling an odd number on a fair six-sided die
odd_numbers <- c(1, 3, 5)
prob_odd <- length(odd_numbers) / 6

# Display probabilities
prob_face_card
prob_odd
```

---

### Axioms of Probability

There are three fundamental axioms of probability:

1. **Non-Negativity**: Probability of any event is ≥ 0.
2. **Additivity**: For mutually exclusive events, the probability of their union is the sum of their probabilities.
3. **Normalization**: The probability of the entire sample space is 1.

**Activity**:  

- **If you roll a die, what is the probability of getting a number less than or equal 3?**
- **What is the probability of rolling a number greater than 4?**

```{r}
# Sample Space
sample_space <- c(1, 2, 3, 4, 5, 6)

# Probability of rolling a number less than 3
prob_less_than_equal_3 <- sum(sample_space <= 3) / length(sample_space)

# Probability of rolling a number greater than 4
prob_greater_than_4 <- sum(sample_space > 4) / length(sample_space)

# Display results
prob_less_than_equal_3
prob_greater_than_4
```

---

## Final Reflection

Now that we've covered some of the major concepts in study design and probability, take a moment to reflect on these questions:

- **What is the most important thing to remember when designing a study?**
- **How can understanding probability help in interpreting real-world situations?**

---