```{r}
#| echo: false
#| message: false
library(tidyverse)
library(ggplot2)
```


# Uniform Distribution

The Uniform distribution is defined by its two parameters, $a$ and $b$, which are the minimum and maximum values, respectively. The pdf of a Uniform distribution is given by:

$f(x) = \frac{1}{b-a} \, \text{ for } x \in [a, b]$

Expected value: $\frac{b + a}{2}$

Variance: $\frac{(b - a)^2}{12}$

```{r}
# Example: Uniform distribution U(0,1)
x <- seq(0, 1, by = 0.01)
y <- dunif(x, min = 0, max = 1)
plot(x, y, type = "l", lwd = 2, ylab = "Density", xlab = "x", main = "PDF of Uniform(0,1)")
```

```{r}
# find the f(0.5)
dunif(x = 0.5, min = 0, max = 1)
```

```{r}
# find the F(0.5)
punif(q = 0.5, min = 0, max = 1)
```

Example:

Imagine a spinner with a continuous circular scale ranging from 0 to 1. 

1. What is the probability that it landing on point 0.5?

2. What is the probability that it landing on a value smaller than 0.5?

# Exponential Distribution

The Exponential distribution is defined by its rate parameter $\lambda$, and is usually used to model the time between events (or time to an event). The pdf of an Exponential distribution is given by:

$f(x) = \lambda e^{-\lambda x} \, \text{ for } x \geq 0$

Expected value: $\frac{1}{\lambda}$

Variance: $\frac{1}{\lambda^2}$

```{r}
# Example: Exponential distribution with lambda = 1
x <- seq(0, 5, by = 0.01)
y <- dexp(x, rate = 1)
plot(x, y, type = "l", lwd = 2, ylab = "Density", xlab = "x", main = "PDF of Exponential(1)")
```

```{r}
# find the f(2)
dexp(x = 2, rate = 1)
```

```{r}
# find the F(2)
pexp(q = 2, rate = 1)
```

Example:

Suppose a bus arrives at a certain bus stop every 20 minutes on average.

1. What is the distribution of the time a person have to wait?

## Insert answer here

2. Find the probability that a bus arrives within 10 minutes

## Insert answer here


If a bus arrives at a certain bus stop every 20 minutes on average, then the rate parameter $\lambda$ is $1/20$ buses per minute.

Now, if we want to find the probability that a bus arrives within 10 minutes, we can substitute these values into the cumulative distribution function (CDF) of the exponential distribution:

$F(x;\lambda) = 1 - e^{-\lambda x}$

Plugging the values, we get:

$F(10; 0.05) = 1 - e^{-(0.05)(10)} = 1 - e^{-0.5} \approx 1 - 0.606 = 0.394$

So, there is approximately a 39.4% chance that a bus will arrive within the next 10 minutes at this bus stop given the new average arrival time of 20 minutes.


# Normal Distribution

The Normal distribution is defined by its mean $\mu$ and variance $\sigma^2$. The pdf of a Normal distribution is given by:

$f(x) = \frac{1}{\sigma \sqrt{2\pi}} e^{-(x-\mu)^2 / (2\sigma^2)}$

Expected value: $\mu$

Variance: $\sigma^2$

```{r}
# Example: Normal distribution with mu = 0 and sigma = 1
x <- seq(-5, 5, by = 0.01)
y <- dnorm(x, mean = 0, sd = 1)
plot(x, y, type = "l", lwd = 2, ylab = "Density", xlab = "x", main = "PDF of Normal(0,1)")
```

```{r}
# find the f(2)
dnorm(x = 3, mean = 0, sd = 1)
```

```{r}
# find the F(2)
pnorm(q = 3, mean = 0, sd = 1)
```

Example: 4.4 on page 160, 4.7 on page 161

## Z-Scores
Z- Scores are used to show how far a certain value is from the mean.
$Z = \frac{X - \mu}{\sigma}$

Essentially, how many standard deviations away a value is from the mean.

Example: Let X represent the heights of adult males and $X \sim N(70, 3^2)$.
What is the Z-score of a male who is 76 inches tall? What does this tell us?


```{r}
mean_height <- 70  
sd_height <- 3     

heights <- seq(60, 80, by = 0.1)

pdf_values <- dnorm(heights, mean = mean_height, sd = sd_height)

height_data <- data.frame(heights, pdf_values)

ggplot(height_data, aes(x = heights, y = pdf_values)) +
  geom_line(linewidth = 1) +
  labs(title = "Normal Distribution of Male Heights",
       x = "Height (inches)",
       y = "Probability Density") 
```



What is the percentile rank of a male who is 76 inches?
```{r}
pnorm(76,70,3)
```


What percentage of men are between the heights 65 inches and 75 inches?

P(65 < X < 75)= P(X < 75)- P(X < 65)
```{r}
pnorm(q=75, mean=70,sd=3)-pnorm(q=65,mean=70,sd=3)
```


If a male is in the 25th percentile, what is the male's height?

```{r}
qnorm(.25, mean=70,sd=3)
```



